---
title: "Seeking Help"
teaching: 20
exercises: 10
questions:
- "what are the basic of R?"
- "How can I get help in R?"
objectives:
- "To be able read R help files for functions and special operators."
- "To be able to use CRAN task views to identify packages to solve a problem."
- "To be able to seek help from your peers."
keypoints:
- "Use `help()` to get online help in R."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

## R has some picky rules

R is an object-oriented language. This means data, information, even functions have to be assigned to an object. This might sounds complicated, but it is really straight forward. Assignment is most done with  the left arrow, `<-`.

Let's create an object with one data point and print it out: 
```{r}
x1 <- 58
x1
```

Object assignment can also be done with the equals sign, `=`:
```{r}
x2 = "hello"
x2
```

But this is much less common among R users.  The most important thing is to **be consistent** with the operator you use. There are occasionally places where it is less confusing to use `<-` than `=`, and it is the most common symbol used in the community. So the recommendation is to use `<-`.


The right hand side of the assignment can be any valid R expression. The right hand side is *fully evaluated* before the assignment occurs.

```{r}
x4 <- x1 * 4
```

Variable names can contain letters, numbers, underscores and periods. They cannot start with a number nor contain spaces at all. Different people use different conventions for long variable names, these include

  * periods.between.words
  * underscores\_between_words
  * camelCaseToSeparateWords

What you use is up to you, but again, **be consistent**.

R is case sensitive! As a commond-line software, each character must be correct (there is no "close enough"). And whitespace matters (sometimes). I'll explain when whitespace matters as it comes up in this workshop                                         `"Yes"` is different from  `"yes"` and  `"Yes   "`

```{r}
# test for equality
"Yes" == "yes"

# count number of characters
nchar("yes")
nchar("yes    ")

# this also illustrates what happens when output is not assigned to an object - it simply gets printed to the console
```
This also illustrated how to include comments in an R file: use `#`. 

R works by calling functions. An R function is like a mathematical function: it takes several arguments and outputs a result or set of results based on what the function is intended to do. The most common format is `function_name(argument1, argument2, ...)`. Each argument is separated by a comma. This is when whitespace is less important. 
```{r}
# here's how to use the function rnorm() to generate a set of normally-distributed numbers given a mean and standard deviation:

x3 <- rnorm(n = 20, mean = 10, sd = 2)
x3
# OR, with whitespace removed: 
rnorm(n=1,mean=10,sd=2)
```

Arguments can be explicitly named, as shown above, or named by expected position. When explicitly assigning variables, always use `=`. The left arrow `<-` does not work properly inside functions. 
Over time, you'll memorize a few functions, or there are always the help files!
```{r}
x4 <- rnorm(20, 10, 2)
```

Some functions don't take any arguments: 

```{r}
#to list all objects in your environment (R session)
ls()
```

You can use `rm` to delete objects you no longer need:

```{r, eval=FALSE}
rm(x1)
rm(x2, x3)
```

## Reading Help files

R, and every package, provide help files for functions. The general syntax to search for help on any function, "function_name", from a specific function that is in a package loaded into your namespace (your interactive R session):

```{r, eval=FALSE}
?function_name
help(function_name)
```

This will load up a help page in RStudio (or as plain text in R by itself).

Each help page is broken down into sections:

 - Description: An extended description of what the function does.
 - Usage: The arguments of the function and their default values.
 - Arguments: An explanation of the data each argument is expecting.
 - Details: Any important details to be aware of.
 - Value: The data the function returns.
 - See Also: Any related functions you might find useful.
 - Examples: Some examples for how to use the function.

Different functions might have different sections, but these are the main ones you should be aware of.

> ## Tip: Reading help files
>
> One of the most daunting aspects of R is the large number of functions
> available. It would be prohibitive, if not impossible to remember the
> correct usage for every function you use. Luckily, the help files
> mean you don't have to!
{: .callout}

## Special Operators

To seek help on special operators, use quotes:

```{r, eval=FALSE}
?"<-"
```

## Getting help on packages

Many packages come with "vignettes": tutorials and extended example documentation.
Without any arguments, `vignette()` will list all vignettes for all installed packages;
`vignette(package="package-name")` will list all available vignettes for
`package-name`, and `vignette("vignette-name")` will open the specified vignette.

If a package doesn't have any vignettes, you can usually find help by typing
`help("package-name")`.

## When you kind of remember the function

If you're not sure what package a function is in, or how it's specifically spelled you can do a fuzzy search:

```{r, eval=FALSE}
??function_name
```

## When you have no idea where to begin

If you don't know what function or package you need to use
[CRAN Task Views](http://cran.at.r-project.org/web/views)
is a specially maintained list of packages grouped into
fields. This can be a good starting point.

## When your code doesn't work: seeking help from your peers

If you're having trouble using a function, 9 times out of 10,
the answers you are seeking have already been answered on
[Stack Overflow](http://stackoverflow.com/). You can search using
the `[r]` tag.

If you can't find the answer, there are a few useful functions to
help you ask a question from your peers:

```{r, eval=FALSE}
?dput
```

Will dump the data you're working with into a format so that it can
be copy and pasted by anyone else into their R session.

```{r}
sessionInfo()
```

Will print out your current version of R, as well as any packages you
have loaded. This can be useful for others to help reproduce and debug
your issue.

> ## Challenge 1
>
> Look at the help for the `c` function. What kind of vector do you
> expect you will create if you evaluate the following:
> ```{r, eval=FALSE}
> c(1, 2, 3)
> c('d', 'e', 'f')
> c(1, 2, 'f')
> ```
> > ## Solution to Challenge 1
> >
> > The `c()` function creates a vector, in which all elements are the
> > same type. In the first case, the elements are numeric, in the
> > second, they are characters, and in the third they are characters:
> > the numeric values are "coerced" to be characters.
> {: .solution}
{: .challenge}

> ## Challenge 2
>
> Look at the help for the `paste` function. You'll need to use this later.
> What is the difference between the `sep` and `collapse` arguments?
>
> > ## Solution to Challenge 2
> >
> > To look at the help for the `paste()` function, use:
> > ```{r, eval=FALSE}
> > help("paste")
> > ?paste
> > ```
> > The difference between `sep` and `collapse` is a little
> > tricky. The `paste` function accepts any number of arguments, each of which
> > can be a vector of any length. The `sep` argument specifies the string
> > used between concatenated terms — by default, a space. The result is a
> > vector as long as the longest argument supplied to `paste`. In contrast,
> > `collapse` specifies that after concatenation the elements are *collapsed*
> > together using the given separator, the result being a single string.
> > e.g. 
> > ```{r} 
> > paste(c("a","b"), "c")
> > paste(c("a","b"), "c", sep = ",")
> > paste(c("a","b"), "c", collapse = "|")
> > paste(c("a","b"), "c", sep = ",", collapse = "|")
> > ```
> > (For more information,
> > scroll to the bottom of the `?paste` help page and look at the
> > examples, or try `example('paste')`.)
> {: .solution}
{: .challenge}

> ## Challenge 3
> Use help to find a function (and its associated parameters) that you could
> use to load data from a csv file in which columns are delimited with "\t"
> (tab) and the decimal point is a "." (period). This check for decimal
> separator is important, especially if you are working with international
> colleagues, because different countries have different conventions for the
> decimal point (i.e. comma vs period).
> hint: use `??csv` to lookup csv-related functions.
> > ## Solution to Challenge 3
> >
> > The standard R function for reading tab-delimited files with a period
> > decimal separator is read.delim(). You can also do this with
> > `read.table(file, sep="\t")` (the period is the *default* decimal
> > separator for `read.table()`, although you may have to change
> > the `comment.char` argument as well if your data file contains
> > hash (#) characters
> {: .solution}
{: .challenge}

## Other ports of call

* [Quick R](http://www.statmethods.net/)
* [RStudio cheat sheets](http://www.rstudio.com/resources/cheatsheets/)
* [Cookbook for R](http://www.cookbook-r.com/)
