---
title: "Challenges - Part 3"
author: "J. Piaskowski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Challenge 1

Write a function `inches_to_mm()` that converts inches to millimeters. There are 2.54 cm in one inch. 

### Challenge 2

Write a function that prints out "hello world!" whenever it is called.
*hint*: use `print()`

### Challenge 3

The Poisson disribution is used frequently in count data, such as occurence of wireworms in an agricultural field. It is defined by one parameter, the mean $\lambda$ (which conveniently, is also the variance). 

```{r fig.width=1.5, fig.height=1}
# library(RCurl)
# script <- getURL("https://raw.githubusercontent.com/rasmusab/distribution_diagrams/master/plot_dist.R", ssl.verifypeer = FALSE)
# eval(parse(text = script))

plot_dist(dists$poisson, labels = c(lambda = expression(lambda)), scale = 1)
```

The predicated number of observations, $\hat{y_i}$, for each discrete count, $k_i$ can be calculated as thus: 

$$\hat{y} = \frac{\lambda^ke^{-\lambda}}{k!}$$
Where k is any real whole number > 0. 

For $\lambda = 7$, write a function to calculate the expected number of observations when k = 1, 2,...,25 (all integers from 1 to 25). *Hint*: use `factorial()` for $k!$.


