---
title: "Challenges - Part 2"
author: "J. Piaskowski"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Challenge 1

We're interested in looking at the sum of the following sequence of fractions:

```
x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
```

This would be tedious to type out, and impossible for high values of n. Use vectorisation to compute x when n=100. What is the sum when n=10,000? 

*Hint*: use `sum()`

## Challenge 2

You have data composed of observations drawn from 2 distributions. We know what distribution each observation belongs to and the parameters of those distributions. 
```{r}
my_data <- as.vector(replicate(5, c(rnorm(1, 10, 2), rnorm(1, 35, 5))))

names(my_data) <- rep(paste0(c("Pop_1", "Pop_2")), 5)

my_data
```

They each need to be mean-centered and scaled so they can be pooled into a shared distribution. The standard way to do that is to substract the mean and divide the difference by the standard deviation (the result are called Z-scores).  The mean and standard deviation for population 1 ("Pop_1") is 10 and 2, respectively. The mean and standard deviation for population 2 ("Pop_2") is 35 and 5, respectively. 

Using vector recycling, obtain the Z-scores for each observation, taking care to use the correct parameters for each distribution. 
*Hint*: first copy and run the code above into your R console so you have the object to work with

